Resilient Landscapes and Livelihoods Through Better Conservation Planning

### 2021-02-26 HCRI_Grant/Proposal.Rmd
updates with cfp letter

---
title: Resilient Landscapes and Livelihoods Through Better Conservation Planning
author: "Matt Clark"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: no
    number_sections: false
  pdf_document:
    toc: no
    number_sections: false
  bookdown::html_document2:
    toc: no
    number_sections: false
fontsize: 12pt
link-citations: yes
csl: pnas.csl
subtitle: Proposal for Boise State Hazard and Climate Resilience Institute Research Grant
bibliography: [Refs.bib]
header-includes:
- \usepackage{pdfpages}
- \usepackage[font={small}]{caption}
---

@@

## Locate figures as close as possible to requested position (=code)
knitr::opts_chunk$set(fig.pos = 'H')
```
\newpage

@@

Results of this project will significantly improve our ability to predict conservation outcomes for both the environment and communities. To communicate simulation outcomes, we will produce an interactive dashboard displaying alternative conservation scenarios. We will host a workshop with our community partner to demonstrate how project results can inform ongoing conservation efforts. This work advances the HCRI Natural Environment pillar by providing a roadmap for conservation planning on Pemba and in complex social-ecological landscapes broadly. This work also contributes to the Economy & Society pillar by explicitly demonstrating the local economic viability of intervention strategies.

@@

\newpage
# Project Narrative
### Background and Rationale
Extreme climatic events such as the hurricane that devastated Puerto Rico in 2017, the deadly European heatwave in 2019, and the recent Texas snowpocalypse are becoming more common due to anthropogenic-caused climate change [@easterling2000climate; @beniston2004extreme].  Intact natural areas buffer the effects of these extreme climatic events on both ecosystems and human communities [@lloret2012extreme; @alongi2008mangrove]. Conserving a broad and diverse array of natural areas is unequivocally our best insurance policy against the worsening impacts of climate change [@isbell2015biodiversity; @loreau2001biodiversity; @oliver2015biodiversity].

Today, on average, we degenerate our natural areas at a rate 56% faster than they regenerate [@network2020global]. In response, $24 billion is allocated to conserving natural areas annually [@waldron2020protecting; @waldron2017reductions]. Thirty six biodiversity hotspots have been identified globally as priority areas for these conservation dollars [@mittermeier2011global; @larsen2012conserving]. These biodiversity hotspots greatly overlap with the world’s poorest communities [@fisher2007poverty]. Because of this overlap and also due to the price and availability of land, many conservation areas are adjacent to, or even contain low income, often indigenous populations [@naughton2005role].

One area in particular, the Coastal Forests of Eastern Africa biodiversity hotspot (Figure \@ref(fig:Mapp)), has seen considerable foregin and community-based investment in conservation [@burgess2017two]. Despite these efforts, deforestation and biodiversity loss in the area continue to worsen [@burgess2017two; @cuadros2018empty]. Pemba Island, our study system for this project is one of two main islands in the Zanzibar archipelago, located in the Coastal Forests of Eastern Africa biodiversity hotspot.

```{r Mapp, echo=FALSE, fig.cap="Pemba Island, Tanzania. Black lines show geographic boundaries for 121 distinct wards on the island. Subfigure shows East Africa with the Coastal Forests of Eastern Africa biodiversity hotspot highlighted in red.", out.width = "40%", fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("InsetSimple.png")
```

Deforestation is a perennial issue on Pemba, as the yearly deforestation rate is nearly 2% and 90% of households rely exclusively on forest products (fuelwood and charcoal) to meet their daily cooking needs [@RGZ2014Zanzibar; @ely2000participatory]. As such, Pemba has been subject to a stream of foregin conservation initiatives since the mid 20th century. Meanwhile, despite the initial enthusiasm for conservation, deforestation remains apace in Pemba. Unfortunately, anecdotal evidence suggests that resentment is mounting amongst residents towards the train of conservation projects that have failed to deliver any kind of sustained economic compensation or ecological benefits.

This phenomenon is not rare. Conservation projects, while well intentioned, often fail to meet both ecological and community development goals [@christie_marine_2004; @watson_performance_2014]. In areas like Pemba, where the environment and livelihoods are closely linked, economic and ecological processes may have complex interactions and feedbacks, which make conservation outcomes difficult to predict.

Community Forests Pemba, our local collaborator, is a Pemban nonprofit organization established in 2009 with a mission to conserve and restore Pembas forests while supporting livelihood development on the island. The organization specifically seeks to build a resilient Pemba, ecologically and economically, with respect to climate change. Community Forests Pemba also works in 35 vulnerable communities across the island to disseminate information on sustainable agroforestry and gender equity to remote households. Our collaborative effort, if funded by this proposal, will seek to build an integrated understanding of why conservation and development initiatives have continued to fall short on Pemba, and how we can do better in the future.

### Aims and Objectives
Forest conservation is a complex mix of individual behaviors, informal group-level institutions, and formal programs or interventions. These social features exist on top of underlying variation in the natural environment. We will construct and parameterize an agent-based land use simulation to produce insights about how all of these moving parts interact and provide a roadmap for implementing successful conservation projects on-the-ground. Our model will generate predictions for ecological and economic viability of two common conservation interventions: protected areas and payments for ecosystem services Table \@ref(tab:tabinterventions).

@@

1. We will create and host an interactive web tool to let individuals run our agent-based simulations to test alternative scenarios and better predict conservation outcomes. This web tool will accompany the scholarly publication of this work to better communicate findings to a broad audience.

2. In the winter following the work outlined here, we will hold a community information and training session with Community Forests Pemba to communicate the findings and implications of this project and to give a live demonstration of the web tool described above. This session will be designed to immediately inform ongoing conservation efforts.

@@

```{r tabinterventions, echo=F, fig.align='center', warning=FALSE}
intVars <- data.frame(Intervention = c("No intervention","Protected areas", "Payments for Ecosystem Services"), 
                      Scheme = c("Individuals harvest renewable resources at a rate which matches their household need, regardless of the location on the landscape or harvest rate compared to resource regeneration rate.",
                                 "Some portion of the landscape is designated for conservation and resources inside are socially unacceptable to harvest. Resource units produced inside conservation area may spillover to use area under sink/source dynamics. Some agents may defect and harvest from the conservation area.",
                                 "Agents are incentivised to forgo harvesting resources in return for economic payments. Payments may offset resource collection when agents previously sold harvested resources, but may fail to change incentive structures when agents directly depend on resources harvested."), 
                      Features = c("N/A",
                                      "Protected area size",
                                      "Market value for harvested fuelwood, Payment amount"))

@@

The broader impacts of this research will advance the overall mission of HCRI to build resilient communities by providing methodology and tools to better plan conservation projects for nature and communities as an integrated system, both on Pemba and elsewhere. The project as described also closely aligns with both the Natural Environment and Economy & Society HRCI pillars in the following ways:

* *Natural Environment* - The work here will provide immediately actionable insights for conservation on Pemba Island and Zanzibar overall. Our community partner, among others, will be provided with the tools to make use of these insights as part of ongoing conservation and community development efforts. More generally, conservation is in need of unifying theory about which kinds of interventions succeed or fail as a result of underlying ecological and social conditions. This research will serve as a foundational building block for conservation science to consider complete social-ecological landscapes moving forward.

@@

### Timeline 
```{r timeline2,echo=FALSE, cache=TRUE, message = FALSE, warning=FALSE,fig.cap="Timeline of proposed research activities and team members involved. Top (green) row shows activities, bottom five (grey) rows show which team members will be involved in each activity.", out.width = "100%", fig.align='center', dpi=72, fig.pos= "H" }

knitr::include_graphics("TimelineBetter.png")
```

@@

### Future Directions
The Human-Environment Systems research center at Boise State has built a partnership with researchers at the Max Planck Institute for Evolutionary Anthropology already working on conservation research on Pemba Island, Tanzania. Although we consider our partnership to be strong, Boise State personnel have not yet visited the field site. If funded, this project will begin what we anticipate will be a long-lasting and fruitful research program. Thus, this proposal will not only fund necessary conservation research, but will also fund an international research partnership between both institutions.

@@

# Matt Clark CV
\includepdf[pages={-}]{CURRICULUMVITAE.pdf}

@@

# Community Partner Letter of Support

>>>>

---
title: Resilient Landscapes and Livelihoods Through Better Conservation Planning
author: "Matt Clark"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: no
    number_sections: false
  pdf_document:
    toc: no
    number_sections: false
  bookdown::html_document2:
    toc: no
    number_sections: false
fontsize: 12pt
link-citations: yes
csl: pnas.csl
subtitle: Proposal for Boise State Hazard and Climate Resilience Institute Research Grant
bibliography: [Refs.bib]
header-includes:
- \usepackage{pdfpages}
- \usepackage[font={small}]{caption}
- \usepackage{wrapfig}
---

@@

## Locate figures as close as possible to requested position (=code)
knitr::opts_chunk$set(fig.pos = 'H')
```

@@

Results of this project will significantly improve our ability to predict conservation outcomes for both the environment and communities. To communicate simulation outcomes, we will produce an interactive dashboard displaying alternative conservation scenarios. We will host a workshop with our community partner to demonstrate how project results can inform ongoing conservation efforts. This work advances the HCRI Natural Environment pillar by providing a roadmap for conservation planning in Pemba and in complex social-ecological landscapes broadly. This work also contributes to the Economy & Society pillar by explicitly demonstrating the local economic viability of intervention strategies.

@@

\newpage
# Project Narrative
### Background and Rationale
Diverse and intact natural landscapes are unequivocally our best insurance policy against the worsening impacts of climate change [@isbell2015biodiversity; @loreau2001biodiversity; @oliver2015biodiversity; @lloret2012extreme; @alongi2008mangrove]. In response, $24 billion is spent on conservation interventions annually [@waldron2020protecting; @waldron2017reductions]. As the world’s most biodiverse areas greatly overlap with the world’s poorest communities, many conservation interventions require community development and buy-in from vulnerable, often indenous populations [@fisher2007poverty; @mittermeier2011global; @larsen2012conserving; @naughton2005role].

\begin{wrapfigure}{r}{0.5\textwidth}
  \begin{center}
    \includegraphics[width=0.48\textwidth]{InsetSimple.png}
  \end{center}
  \caption{Pemba Island, Tanzania. Black lines show geographic boundaries for 121 distinct wards on the island. Subfigure shows East Africa with the Coastal Forests of Eastern Africa biodiversity hotspot highlighted in red.}
\end{wrapfigure}

Conservation interventions, while well intentioned, frequently fail to meet both ecological and community development goals [@christie_marine_2004; @watson_performance_2014]. When environments and livelihoods are closely linked, economic and ecological processes may have complex interactions and feedbacks that make conservation outcomes difficult to predict. One area in particular, the Coastal Forests of Eastern Africa biodiversity hotspot, has seen considerable foreign and community-based investment in conservation, yet deforestation and biodiversity loss in the area continue to worsen [ @burgess2017two; @cuadros2018empty].

This study focuses on Pemba Island, Tanzania, located in the Coastal Forests of Eastern Africa biodiversity hotspot. Deforestation is a perennial issue in Pemba, as the yearly deforestation rate is nearly 2%, and 90% of households rely exclusively on forest products (fuelwood and charcoal) to meet their daily cooking needs [@RGZ2014Zanzibar; @ely2000participatory]. As such, Pemba has been subject to a stream of foreign conservation initiatives since the mid 20th century, yet deforestation remains apace. Unfortunately, anecdotal evidence suggests that resentment is mounting amongst residents towards the train of conservation projects that have failed to deliver any sustained economic compensation or ecological benefits. Our collaborative effort will build land use simulations that integrate ecology and economics to demonstrate why conservation and development initiatives have continued to fall short in Pemba, and how we can do better in the future.

This project is in collaboration with the Max Planck Institute for Evolutionary Anthropology, and Community Forests Pemba, a Pemban nonprofit organization established in 2009 with a mission to conserve and restore Pemba’s forests while supporting livelihood development on the island. The organization specifically seeks to build a resilient Pemba, ecologically and economically, with respect to climate change. This effort is mutually beneficial as Community Forests Pemba facilitate data collection and provide on-the-ground logistical support, while resulting land use simulations will inform future conservation efforts by the organization

If funded, the project will establish a formal collaboration between the Hazards and Climate Resilience Center, the Max Planck Institute for Evolutionary Anthropology, Community Forests Pemba, and Boise State Human-Environment Systems. This funding will specifically support the dissertation work of one PhD student at Boise State, Matt Clark, and will yield an interactive web dashboard displaying all land use simulations described below.

```{r Mapp, echo=FALSE,eval=FALSE, fig.cap="Pemba Island, Tanzania. Black lines show geographic boundaries for 121 distinct wards on the island. Subfigure shows East Africa with the Coastal Forests of Eastern Africa biodiversity hotspot highlighted in red.", out.width = "40%" ,fig.align='center', dpi=72, fig.pos= "H"}
knitr::include_graphics("InsetSimple.png")
```

### Aims and Objectives
Our project will generate predictions for ecological and economic viability of two common conservation interventions: protected areas and payments for ecosystem services (Table \@ref(tab:tabinterventions), and provide a roadmap for implementing successful conservation projects on-the-ground.

@@

1. We will create and host an interactive web dashboard to let individuals run our agent-based simulations to visualize alternative scenarios and better predict conservation outcomes. This web tool will accompany the scholarly publication of this work to better communicate findings to a broad audience.

2. We will hold a community information and training session with Community Forests Pemba to communicate the findings and implications of this project and to give a live demonstration of the web dashboard described above. This session will be designed to immediately inform ongoing conservation efforts.

@@

All simulations will be displayed visually on an interactive dashboard created using the Flexdashboard and R Shiny functionality in the R statistical programming language [@R; @flexD; @shiny]. The dashboard will allow users to test all three intervention scenarios and their variations against all described social and ecological starting conditions.

```{r tabinterventions, echo=F, fig.align='center', warning=FALSE}
intVars <- data.frame(Intervention = c("No intervention","Protected areas", "Payments for ecosystem services"), 
                      Scheme = c("Individuals harvest renewable resources at a rate which matches their household need, regardless of the location on the landscape or harvest rate compared to resource regeneration rate.",
                                 "Some portion of the landscape is designated for conservation and resources inside are socially unacceptable to harvest. Resource units produced inside conservation area may spillover to use area under sink/source dynamics. Some agents may defect and harvest from the conservation area.",
                                 "Agents are incentivised to forgo harvesting resources in return for economic payments. Payments may offset resource collection when agents previously sold harvested resources, but may fail to change incentive structures when agents directly depend on resources harvested."), 
                      Features = c("N/A",
                                      "Protected area size",
                                      "Market value for harvested fuelwood, Payment amount"))

@@

The broader impacts of this research will advance the overall mission of HCRI to build resilient communities by providing methodology and tools to better plan conservation projects for nature and communities as an integrated system, both in Pemba and elsewhere. The project as described also closely aligns with both the Natural Environment and Economy & Society HRCI pillars in the following ways:

* *Natural Environment* - The work here will provide immediately actionable insights for conservation in Pemba Island and Zanzibar overall. Our community partner, among others, will be provided with the tools to make use of these insights as part of ongoing conservation and community development efforts. More generally, conservation is in need of unifying theory about which kinds of interventions succeed or fail as a result of underlying ecological and social conditions. This research will serve as a foundational building block for conservation science to consider complete social-ecological landscapes moving forward.

@@

### Timeline 
```{r timeline2,echo=FALSE, cache=TRUE, message = FALSE, warning=FALSE,fig.cap="Timeline of proposed research activities and team members involved.", out.width = "80%", fig.align='center', dpi=72, fig.pos= "H" }

knitr::include_graphics("Timeline.png")
```

@@

### Future Directions
The Human-Environment Systems research center at Boise State has built a partnership with researchers at the Max Planck Institute for Evolutionary Anthropology already working on conservation research in Pemba Island, Tanzania. Although we consider our partnership to be strong, Boise State personnel have not yet visited the field site. If funded, this project will begin what we anticipate will be a long-lasting and fruitful research program. Thus, this proposal will not only fund necessary conservation research, but will also fund an international research partnership between both institutions.

@@

\includepdf[pages={-}]{CURRICULUMVITAE.pdf}

@@

\includepdf[pages={-}]{CFP_Letter.pdf}
